---
pagetitle: "Introduction to Statistical Computing - dplyr"
abstract: |
 This is the third of a five part series of lectures as part of the data handling postgraduate module in Atlantic Technological University. 
author: "Eugene Hickey"
date: "May 2023"
host: "Sligo Campus"
institute: "Atlantic Technological University"
strip-comments: true
format: 
  revealjs:
    transition: slide
    transition-speed: fast
    logo: images/atu-logo.png
    footer: "[Statistical Computing](https://vlesligo.atu.ie)"
    # embed-resources: true
    slide-number: c/t
    multiplex: true
    theme: solarized
    show-slide-number: all
    controls: true
    width: 1280
    height: 720
    incremental: false
    auto-stretch: false
    css: [assets/custom.css, assets/tachyons-addon.css]
    include-after: |
      <link rel="stylesheet" href="assets/syntax-highlight.css">
      <link rel="stylesheet" href="assets/fontawesome-free-6.1.1-web/css/all.min.css">
      <script src="assets/fontawesome-free-6.1.1-web/js/all.min.js"></script>
callout-icon: false
output-file: index.html
execute:
  echo: true
  message: false
  warning: false
---

##  {#title-slide background="images/dawn.JPG"}

```{r setup, include = FALSE}
library(tidyverse)
library(showtext)

rotating_text <- function(x, align = "top") {
  glue::glue('
<div style="overflow: hidden; height: 1.2em;">
<ul class="content__container__list {align}" style="text-align: {align}">
<li class="content__item">{x[1]}</li>
<li class="content__item">{x[2]}</li>
<li class="content__item">{x[3]}</li>
<li class="content__item">{x[4]}</li>
</ul>
</div>')
}

fa_list <- function(x, incremental = FALSE) {
  icons <- names(x)
  fragment <- ifelse(incremental, "fragment", "")
  items <- glue::glue('<li class="{fragment}"><span class="fa-li"><i class="{icons}"></i></span> {x}</li>')
  paste('<ul class="fa-ul">', 
        paste(items, collapse = "\n"),
        "</ul>", sep = "\n")
}

```

```{r ggplot-theme}
#| cache: true
#| echo: false

my_font <- "Neucha"
my_font <- "Coming Soon"
font_add(family = my_font, regular = "assets/ComingSoon-Regular.ttf")
showtext_auto()
theme_clean <- function() {
    theme_minimal(base_family = my_font) +
        theme(panel.grid.minor = element_blank(),
              text = element_text(size = 32, family = my_font),
              plot.background = element_rect(fill = "white", color = NA),
              axis.text = element_text(size = 32),
              axis.title = element_text(face = "bold", size = 28),
              strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0),
              strip.background = element_rect(fill = "grey80", color = NA),
              legend.text = element_text(size = 36))
}
```


::: title-box
<h2>`r rmarkdown::metadata$pagetitle`</h2>

👨‍💻 [`r rmarkdown::metadata$author` \@ `r rmarkdown::metadata$institute`]{.author} 👨‍💻

`r rotating_text(c('<i class="fa-solid fa-envelope"></i> eugene.hickey@tudublin.ie', '<i class="fa-brands fa-mastodon"></i> @eugene100hickey', '<i class="fa-brands fa-github"></i> github.com/eugene100hickey', '<i class="fa-solid fa-globe"></i> www.fizzics.ie'))`
:::

------------------------------------------------------------------------

##  Week Four - Statistical Inference

::: {style="font-size: 100%;"}

- Time we did some statistics on our data

- This ties in nicely with John's half of our course

- We'll look at this from three perspectives

    - distributions
    
    - confidence intervals
    
    - hypothesis tests

:::

## Distributions

- There are lots of these, but we'll focus on four

    - binomial
    
    - poisson
    
    - normal
    
    - exponential
    
    
## Binomial Distribution

- number of _successes_ in a sequence of trials

- three parameters

    - the number of trials, **n**
    
    - the number of successes, **k**
    
    - the probability of success in a single trial, **p**
    
---

$$ {\displaystyle f(k,n,p)=\Pr(k;n,p)=\Pr(X=k)={\binom {n}{k}}p^{k}(1-p)^{n-k}}$$
for k = 0, 1, 2, ..., n, where


$$ {\displaystyle {\binom {n}{k}}={\frac {n!}{k!(n-k)!}}} $$

## What this looks like:

```{r binom-setup}
#| echo: false

n <- 100
p <- 0.85
```


A rocket launch has a probability of success of `r p`%. From a `r n` launches, what is the profile of the number of successes?

```{r binom-1}
#| echo: false

tibble(x = 1:n, y = dbinom(1:n, n, p)) |> 
  ggplot(aes(x, y))  + 
  geom_point(size = 5, colour = "firebrick4") + 
  geom_line(alpha = 0.5, colour = "firebrick4") +
  labs(title = "Number of successful rocket launches", y = "", x = "Successful Launches") + 
  theme_clean() +
  theme(plot.title.position = "plot")

```

## How this works in R

```{r}
dbinom(x = 90, size = 100, prob = 0.85)
```

```{r}
dbinom(x = 85, size = 100, prob = 0.85)
```

```{r}
dbinom(x = 80, size = 100, prob = 0.85)
```

```{r}
dbinom(x = 70, size = 100, prob = 0.85)
```

```{r}
dbinom(x = 60, size = 100, prob = 0.85)
```

```{r}
dbinom(x = 50, size = 100, prob = 0.85)
```


## `dbinom()`, `pbinom()`, `qbinom()`, and `rbinom()`

::: {style="font-size: 80%;"}

- `dbinom()` calculates the probability of a specific outcome value

- `pbinom()` calculates the sum of probabilities less than (or greater than), an outcome value

    - this is the _cumulative probability_
    
    - adds up a whole bunch of `dbinoms()`
    
- `qbinom()` calculates the number of successes below which there is a certain probability

    - like the inverse function of `pbinom()`
    
    - give it a probability of it works out the number of successes threshold
    
- `rbinom()` generates random numbers of _successs_ from a binomial distribution

:::

## `pbinom()`

::::{.columns}
:::{.column width="50%"}

```{r}
#| echo: false

theme_set(
  theme_clean() +
  theme(plot.title.position = "plot"))
```


```{r}
#| eval: false
pbinom(1:100,100,0.85,lower.tail=TRUE)
```


```{r binom-2}
#| echo: false

tibble(x = 1:n, y = pbinom(1:n, n, p, lower.tail=T)) |> 
  ggplot(aes(x, y))  + 
  geom_point(size = 5, colour = "firebrick4") + 
  geom_line(alpha = 0.5, colour = "firebrick4") +
  labs(title = "Probability of getting this or less than this\nnumber of successful rocket launches", y = "", x = "Successful Launches")

```

:::
:::{.column width="50%"}

```{r}
#| eval: false
pbinom(1:100,100,0.85,lower.tail=FALSE)
```


```{r binom-3}
#| echo: false

tibble(x = 1:n, y = pbinom(1:n, n, p, lower.tail = F)) |> 
  ggplot(aes(x, y))  + 
  geom_point(size = 5, colour = "firebrick4") + 
  geom_line(alpha = 0.5, colour = "firebrick4") +
  labs(title = "Probability of getting more than this\nnumber of successful rocket launches", y = "", x = "Successful Launches")

```

:::
::::

::: {style="font-size: 80%;"}

- `lower.tail = TRUE` means _k_ successes or less (successes $\le$ threshold, k)
 
- `lower.tail = FALSE` means more than _k_ successes (successes $>$ threshold, k)

:::

## Example

- Family of eight children, chance of getting at least 7 girls

```{r}
dbinom(x = 7, size = 8, prob = 0.5) + dbinom(x = 8, size = 8, prob = 0.5)
```

```{r}
pbinom(q = 6, size = 8, prob = 0.5, lower.tail = FALSE)
```

Note we had to use _k - 1 = 6_ in `pbinom()` calculation
