```{r data generation, echo = FALSE, results = "hide", message=F, warning=F}
library(exams)
library(tidyverse)
library(gt)
library(showtext)

font_add_google(name = "Caveat", family = "Caveat")
showtext_auto()

z <- read_rds("https://github.com/eugene100hickey/ATU-2023/blob/main/rexams/data/ABAData-wide?raw=true")

genes <- colnames(z)[2:dim(z)[2]] |> 
  sample(10)


z1 <- z |> 
  select(structure, any_of(genes)) |> 
  pivot_longer(-structure, names_to = "gene", values_to = "signal")

solution <- z1$signal |> signif(3)
```


Question
========
Run the code line below to read in a data frame called `brain_gene_expression_wide`. 

brain_gene_expression_wide <- read_rds("https://github.com/eugene100hickey/ATU-2023/blob/main/rexams/data/ABAData-wide?raw=true") |> 
  select(structure, `r noquote(paste(genes, ","))`)
  
(note that if you cut and paste in this text, you might have to rewrite the quotation marks, __"__)


Use `pivot_long()` torecast this data in long format. The top of this table is shown below. Use this to calculate the mean level of gene expression. 

```{r echo=FALSE}
z1 %>% 
  head() |> 
  gt() %>% 
  cols_label(structure = "Brain Area Code",
             gene = md("Gene Name"), 
             signal = md("Gene Expression")) %>% 
  cols_align(align = "center") %>% 
  tab_options(column_labels.background.color = "cornflowerblue", 
              table.font.names = "Caveat",
              table.font.style = "oblique",
              column_labels.font.weight = "bold",
              column_labels.vlines.width = 1,
              column_labels.vlines.color = "darkblue",
              column_labels.vlines.style = "solid",
              table_body.vlines.width = 1,
              table_body.vlines.color = "darkblue",
              table_body.vlines.style = "solid"
              )
```


Solution
========
Use `pivot_longer(-structure, names_to = "gene", values_to = "signal")`.



Meta-information
================
extype: num
exsolution: `r solution`
exname: ABAData-pivot-long
extol: `r solution*0.02`
